The zoon R package for reproducible and shareable species distribution modelling

### 2016-12-20 ms.Rmd
polishing module section

# The zoon R package

@@

#### Module types

The zoon R package encodes SDM analyses as a simple workflow of five key steps, with each step encoded as one or more software modules. Users combine these modules via a call to the `workflow` function, which executes each module in turn, before returning a `zoonWorkflow` object - a shareable, extensible and fully reproducible documentation of the SDM analysis. Figure 1 illustrates the structure of a zoon workflow comprising the five module types and illustrates how multiple modules may be combined at each stage. The module types and module combination options are described in more detail below.

```{r flow_diagram, echo = FALSE, out.width = "375px", fig.cap = "The modular SDM structure encoded by a zoon workflow. A) Flow diagram representing the five module types and intermediate objects: df - an occurrence dataframe; ras - a RasterStack object; mod - a ZoonModel object; ? - any user-defined inputs or outputs. B) The flow diagram implied by chaining modules of the same type. C) The flow diagram implied by listing modules of the same type. Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the zoon vignette 'Building a module'."}
knitr::include_graphics("figs/diagrams.png")
```

#### Lists, Chains and Replicates

Many SDM analyses apply more than one method for each of these steps.
zoon enables more complex workflows by enabling users to pass multiple modules of each type via the List, Chain and Replicate functions.

To combine multiple modules of the same type we provide the `Chain` command.
For occurrence and covariate modules, this command takes multiple modules and simply combines the data acquired by each module.
Chained process models are run sequentially.
For example if a user wants to generate background or pseudo absence data and then split the data into crossvalidation folds, modules implementing these two seperate process would be chained in that order.
Finally, chained output modules are simply all run seperately allowing the user to create multiple maps and summary figures, calcualte performance metrics and create other model outputs in one workflow.
Model modules cannot be chained.

Many SDM analyses require running similar workflow in parallel. 
In applied settings this may be running the same analysis for multiple species.
In methodological work, it is important to compare new methods or models to benchmarks.
To divide a zoon workflow into multiple parallel analyses the `list` function is used.
`list` can be used to split the analysis at any point: listing multiple occurrence modules together will run the workflow on different occurrence datasets while listing multiple model modules will fit different models to a shared, identical dataset but then apply output modules to each model separately. In cases where modules containing randomness need to be run many times - such as running a workflow with hundreds of simulated datasets - `replicate` can be used. This simply creates a `list` of the specified module, repeated a given number of times.

The following workflow demonstrates use of the `Chain`s and `list`s to fit three different SDMs to a presence-background dataset for a UK mosquito species, after standardising the covariates. The results of the three models are then returned in Shiny app allowing the user to interactively explore the occurrence and covariate data, model summaries and prediction maps.

@@

### Inspecting, sharing and extending a zoon workflow

@@

To do this, it first is necessary to provide additional metadata including the module tile and description, module type, author name and definitions for any non-mandatory arguments.
zoon provides the function `BuildModule()` to facilitate entering this metadata and checking that it matches the module.

@@

To convert this code into a zoon module, we need to write it to a standalone file (named `spThin.R`) with the metadata required to build the module documentation. The zoon function `BuildModule()` helps with this step, and can also run checks to make sure we got everything right:

>>>>

@@

# Building a workflow

zoon encodes SDM analyses as a simple workflow of five key steps: obtaining occurrence data; obtaining covariate data; applying some processes to these data; fitting one or more models; and generating any outputs (Figure 1A).
Each of these steps is encoded as one or more software modules of a given type.
Users combine these modules via a call to the `workflow` function, which executes each module in turn, before returning a `zoonWorkflow` object - a shareable, extensible and fully reproducible documentation of the SDM analysis.
Figure 1B illustrates how data and outputs are passed between modules of the five different types in a a zoon workflow.

```{r flow_diagram, echo = FALSE, out.width = "375px", fig.cap = "The modular SDM structure encoded by a zoon workflow. A) Description of the five module types. B) Flow diagram illustrating how objects are passed between different module types: `data frame' - an occurrence dataframe; `ras' - a RasterStack object; `mod' - a ZoonModel object; `?' - any user-defined inputs or outputs. C) The flow diagram implied by chaining two `process' modules. D) The flow diagram implied by listing three `model' modules. Full details of module inputs and outputs, and the effects of listing and chaining each module type are given in the zoon vignette `Building a module'."}
knitr::include_graphics("figs/diagrams.png")
```

## Lists, Chains and Replicates

Many SDM analyses apply more than one method in one of the five SDM steps, for example to combine multiple covariates and processing steps, to compare various models on the same dataset, or to run the same procedure for several species.
zoon enables these more complex workflows by enabling users to pass multiple modules of each type via the `list`, `Chain` and `Replicate` functions.

### Chain

`Chain` runs multiple modules of the same type *sequentially*, as illustrated in Figure 1C.
For example, in a presence-only analysis a chain of two process modules could be used to generate background data and then to standardise the covariate rasters using the presence/background dataset.
Chains can also be applied to occurrence or covariate modules, to combine multiple datasets (e.g. occurrences from different databases) into one.
Chaining output modules simply runs each module separately, allowing the user to create multiple maps and summary figures, calcualte performance metrics and create other model outputs in one workflow.
Model modules are the only module type that may not be chained, since their outputs and inputs are always different.

### list

`list` splits a workflow, with each path using a different one of the listed modules, as illustrated in Figure 1D.
For example listing multiple model modules would take the same occurrence and covariate data and fit three separate models, applying the same output modules to each model separately.
Lists can also be used to run the same SDM procedure for multiple species, or for multiple sets of covariates, or to compare different process modules.
Listing output modules has the same effect as chaining them, each module is applied separately.

### Replicate

Some steps in SDM analyses are stochastic and need to be run multiple times; `Replicate` enables this by generating a list with one module repeated a given number of times.
`Replicate` could be used to run the same workflow for hundreds of simulated occurrence or covariate datasets, to generate multiple bootstraps for modelling, or to fit models with stochastic elements

### Example workflow

The following workflow demonstrates use of the chains and lists to fit three different SDMs to a presence-background dataset for a UK mosquito species, after standardising the covariates.
The results of the three models are then returned in Shiny app allowing the user to interactively explore the occurrence and covariate data, model summaries and prediction maps.

@@

## Inspecting, sharing and extending a zoon workflow

@@

To do this, it first is necessary to provide additional metadata including the module tile and description, module type, author name and definitions for any non-mandatory arguments.
zoon provides the function `BuildModule` to facilitate entering this metadata and checking that it matches the module.

@@

To convert this code into a zoon module, we need to write it to a standalone file (named `spThin.R`) with the metadata required to build the module documentation. The zoon function `BuildModule` helps with this step, and can also run checks to make sure we got everything right:
